``` {r housekeeping, include = FALSE}

setwd("C:\\Users\\micha\\Documents\\JohnsHopkins\\JohnsHopkins_Course5_ReproducibleResearch\\Week 2\\RepData_PeerAssessment1")

```

---
title: "PA1_template"
author: "Michael McGill"
date: "8/7/2020"
output: html_document
---

## Loading required packages
  
``` {r loadpackages, message = FALSE}

library(lubridate)
library(dplyr)
library(lattice)

```

## Loading and preprocessing the data
  
Download data and load into R

``` {r loaddata}

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
zip <- ".\\repdata_data_activity.zip"
file <- ".\\activity.csv"

if (!file.exists(zip)) download.file(url, zip)
if (!file.exists(file)) unzip(zip)

df <- read.csv(file, colClasses = c("numeric", "character", "character"))

```

Convert the date variable to a date format

``` {r dateformat}

df$date <- ymd(df$date)

```

Create a time variable

``` {r timeformat}

minutes <- substr(df$interval, nchar(df$interval) - 1, nchar(df$interval))
hours <- substr(df$interval, 1, nchar(df$interval) - 2)
hours <- sapply(hours, function(x) if(nchar(x) == 0) 0 else x)

df$time <- paste(hours, ":", minutes, sep = "") %>% hm() %>% as_date()

```

## What is mean total number of steps taken per day?
  
Calculate the total number of steps taken per day

``` {r sumdailysteps, message = FALSE}

dailysum_df <- df %>% group_by(date) %>% 
               summarise(dailysum = sum(steps, na.rm = TRUE))

```
  
Make a histogram of the total number of steps taken each day

``` {r histsteps}

hist(dailysum_df$dailysum, xlab = "Number of steps", main = NULL)

```

Calculate and report the mean and median of the total number of steps taken per day

``` {r meansteps}

mean(dailysum_df$dailysum, na.rm = TRUE)


```

``` {r medsteps}

median(dailysum_df$dailysum, na.rm = TRUE)

```

## What is the average daily activity pattern?
  
Make a time series plot of the 5-minute interval (x-axis) and the average number
of steps taken, averaged across all days (y-axis)

``` {r avgintervalsteps, message = FALSE}

intervalavg_df <- df %>% group_by(time) %>%
        summarise(intervalavg = mean(steps, na.rm = TRUE))

plot(intervalavg_df$time, intervalavg_df$intervalavg, type = "l",
     xlab = "Interval", ylab = "Number of steps",)

```
  
Which 5-minute interval, on average across all the days in the dataset,
contains the maximum number of steps?

``` {r maxinterval}

intervalavg_df[[which.max(intervalavg_df$intervalavg),1]] %>% format("%H:%M")

```

## Imputing missing values

Calculate and report the total number of missing values in the dataset
(i.e. the total number of rows with NAs)

``` {r numna}

sum(is.na(df$steps))

```
Create a new dataset that is equal to the original dataset but with the
missing data filled in. NAs will be replaced with the mean for that 5-minute
interval.

``` {r nonas}

df_nonas <- df %>% group_by(time) %>%
                   mutate(steps = ifelse(is.na(steps),
                                         mean(steps, na.rm = TRUE),
                                         steps))

```

Calculate the total number of steps taken per day

``` {r sumdailysteps_nonas, message = FALSE}

dailysum_df_nonas <- df_nonas %>% group_by(date) %>% 
                     summarise(dailysum = sum(steps, na.rm = TRUE))

```

Make a histogram of the total number of steps taken each day.

``` {r histsteps_nonas}

hist(dailysum_df_nonas$dailysum, xlab = "Number of steps", main = NULL)

```

Calculate and report the mean and median of the total number of steps taken per day

``` {r meansteps_nonas}

mean(dailysum_df_nonas$dailysum, na.rm = TRUE)


```

``` {r medsteps_nonas}

median(dailysum_df_nonas$dailysum, na.rm = TRUE)

```

## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels 
– “weekday” and “weekend” indicating whether a given date is a
weekday or weekend day.

``` {r isweekday}

df_nonas$weekday <- ifelse(wday(df_nonas$date) != 6 & wday(df_nonas$date) != 7,
                           "weekday", "weekend")

df_nonas$weekday <- as.factor(df_nonas$weekday)

```

Make a panel plot containing a time series plot of the 5-minute interval
(x-axis) and the average number of steps taken, averaged across all weekday days
or weekend days (y-axis).

``` {r avgintervalsteps_nonas, message = FALSE}

intervalavg_df_nonas <- df_nonas %>% group_by(weekday, time) %>%
                        summarise(intervalavg = mean(steps, na.rm = TRUE))

xlabs <- c("00:00", "05:00", "10:00", "15:00", "20:00") %>%
         hm() %>% as_date()

xyplot(intervalavg ~ time | weekday, intervalavg_df_nonas, 
       type = "l", layout = c(1,2),
       xlab = "Interval", ylab = "Number of steps",
       scales = list(x = list(at = xlabs,
                              labels = xlabs %>% format("%H:%M"))))

```